---
title: "Getting started with CPLASS"
author:
  - "Linh Do"
  - "Scott A. McKinley"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting started with CPLASS}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
set.seed(20260706)
```

## Overview

**CPLASS** (Continuous Piecewise-Linear Approximation via Stochastic Search)
detects changes in velocity within multidimensional particle trajectories.
Unlike traditional changepoint methods that target changes in *mean*, CPLASS
models a *continuous* piecewise-linear anchor trajectory and searches the
space of changepoint configurations with a tailored Metropolis-Hastings
sampler. See Do et al. (2026), *Change-in-velocity detection in
multidimensional data*, for the full statistical model and simulation
studies behind the method.

This vignette walks through:

1. loading the package and an example trajectory data set;
2. running CPLASS on a single path;
3. visualizing the inferred continuous piecewise-linear anchor path;
4. checking whether the search has stabilized, and running multiple
   independent starts for a more reliable fit;
5. summarizing a fitted trajectory with the Cumulative Speed Allocation
   (CSA); and
6. running CPLASS on a small collection of paths at once.

CPLASS is a stochastic *search* for the configuration that maximizes a
penalized criterion, not a Bayesian sampler in the usual sense. So
"convergence" below always means **optimization has stabilized** (no better
configuration is being found), not that a Markov chain has reached a
stationary posterior distribution.

## 1. Load the package

```{r load-libraries}
library(cplass)
library(ggplot2)
library(dplyr)
library(patchwork)
```

```{r package-check}
packageVersion("cplass")
```

## 2. Load the example trajectory data

The package ships a small example data set of lysosome trajectories,
collected in the peripheral region of cells (Rayens et al. 2022). Each row
is the position of one tracked particle at one observed time point.

| Column        | Description                          |
|:--------------|:--------------------------------------|
| `index_path`  | Trajectory identifier                 |
| `t`           | Observed time, in seconds             |
| `x`           | x-position of the particle ($\mu$m)   |
| `y`           | y-position of the particle ($\mu$m)   |

```{r load-data}
data_file <- system.file("extdata", "Real_21_Periphery.csv", package = "cplass")
traj_data <- read.csv(data_file)
str(traj_data)
```

Number of observations per path:

```{r data-summary}
path_summary <- traj_data |>
  group_by(index_path) |>
  summarise(
    n_obs = n(),
    duration = max(t) - min(t),
    .groups = "drop"
  )
head(path_summary)
```

## 3. Inspect one trajectory

We use `index_path == 3` as the running example.

```{r choose-path}
path_id <- 3
path <- traj_data |> filter(index_path == path_id)

t <- path$t
x <- path$x
y <- path$y
```

```{r raw-path-plot, fig.height=7}
p_xy <- ggplot(path, aes(x = x, y = y)) +
  geom_path() +
  geom_point(size = 0.8, alpha = 0.7) +
  coord_equal() +
  theme_minimal() +
  labs(title = paste("Raw trajectory: path", path_id), x = "x", y = "y")

p_xt <- ggplot(path, aes(x = t, y = x)) +
  geom_line() +
  theme_minimal() +
  labs(title = "x-coordinate over time", x = "time (s)", y = "x")

p_yt <- ggplot(path, aes(x = t, y = y)) +
  geom_line() +
  theme_minimal() +
  labs(title = "y-coordinate over time", x = "time (s)", y = "y")

p_xy / p_xt / p_yt
```

## 4. Run CPLASS on a single path

The main inputs are the observed time vector `t` and the coordinate vectors
`x`, `y`. Useful tuning parameters:

* `iter_max` -- total number of stochastic-search iterations.
* `burn_in` -- initial iterations discarded before selecting the best
  configuration.
* `s_cap` -- speed above which the (optional) speed penalty is applied.
* `speed_pen` -- whether to use the speed penalty (Section 2.3 of the paper).
* `gamma` -- exponent in the strengthened SIC penalty (paper recommends 1.01).
* `pen` -- penalty choice: `"ssic"`, `"aicc"`, or `"hybrid"`.
* `patience` -- if set, stop early once the criterion hasn't improved for
  this many iterations (see `?CPLASS`).

```{r run-single-cplass, results='hide'}
fit_single <- CPLASS(
  t, x, y,
  iter_max = 5000,
  burn_in = 500,
  s_cap = 1,
  gamma = 1.01,
  speed_pen = TRUE,
  pen = "ssic",
  show_progress = FALSE
)
```

The fit is a list with a tibble of inferred segments and the reconstructed
anchor path:

```{r single-output-summary}
names(fit_single)
fit_single$segments_inferred
```

```{r segment-summary}
fit_single$segments_inferred |>
  summarise(
    n_segments = n(),
    total_duration = sum(durations),
    mean_speed = mean(speeds),
    max_speed = max(speeds),
    prop_motile_time = sum(durations[states == 1]) / sum(durations)
  )
```

## 5. Visualize the fit

The black line is the observed trajectory; the blue line is the inferred
continuous piecewise-linear anchor path. Shaded regions mark stationary vs.
motile segments using the default speed cutoff of 0.1 $\mu$m/s.

```{r visualize-single-fit}
max_speed <- ceiling(max(fit_single$segments_inferred$speeds))

plot_path_inferred(
  fit_single,
  motor = "Perinuclear lysosome",
  max_speed = max_speed,
  t_lim = c(0, ceiling(max(t) - min(t))),
  path_index = path_id
)
```

## 6. Has the search stabilized?

`check_convergence()` re-runs CPLASS with its diagnostic trace enabled and
checks how long it has been since the running-maximum criterion score last
improved. This is useful for sizing `iter_max` for trajectories of a given
length -- see Figure 4 of the companion paper for why a naive gradient-based
stopping rule is not reliable here.

```{r check-convergence, results='hide'}
conv <- check_convergence(t, x, y, iter_max = 5000, burn_in = 500, show_progress = FALSE)
```

```{r convergence-summary}
conv$converged
conv$best_score
conv$last_improvement_iter
```

```{r plot-convergence}
plot_convergence(conv)
```

## 7. Multiple independent starts

The penalized criterion can have several local maxima (Figure 4 of the
paper). `CPLASS_multistart()` runs several independent chains and keeps the
one with the highest criterion score, using patience-based early stopping
by default so the extra starts don't cost a full `iter_max` each.

```{r multistart}
ms <- CPLASS_multistart(
  t, x, y,
  iter_max = 5000,
  burn_in = 500,
  n_starts = 5,
  n_cores = 1,
  show_progress = FALSE
)

ms$all_scores
ms$score_spread
```

A small `score_spread` means the starts agree on (approximately) the same
optimum. Use the winning run for downstream analysis:

```{r multistart-winning-fit}
fit_best <- ms$best
fit_best$segments_inferred
```

## 8. Cumulative Speed Allocation (CSA)

The CSA is a duration-weighted empirical distribution of inferred segment
speeds (Cook et al. 2025): for each speed threshold, the proportion of time
the particle spent moving slower than that threshold. It summarizes an
entire trajectory's speed profile without an arbitrary motile/stationary
cutoff.

```{r csa-single}
speed_mesh <- seq(0, max(fit_best$segments_inferred$speeds), length.out = 100)
csa_single <- compute_csa(fit_best$segments_inferred, speed_mesh = speed_mesh)
head(csa_single)
```

```{r plot-csa-single}
ggplot(csa_single, aes(x = s, y = csa)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    title = paste("CSA curve: path", path_id),
    x = "Speed threshold",
    y = "Duration-weighted proportion of time"
  )
```

## 9. Running CPLASS on a collection of paths

For a batch of trajectories, loop over paths and collect the fits. For a
full analysis, prefer `CPLASS_multistart()` per path over a single
`CPLASS()` call.

```{r run-multiple-paths, results='hide'}
path_ids <- sort(unique(traj_data$index_path))[1:3]

fit_list <- vector("list", length(path_ids))
names(fit_list) <- paste0("path_", path_ids)

for (ii in seq_along(path_ids)) {
  id <- path_ids[ii]
  path_i <- traj_data |> filter(index_path == id)
  fit_list[[ii]] <- CPLASS(
    path_i$t, path_i$x, path_i$y,
    iter_max = 5000, burn_in = 500, show_progress = FALSE
  )
}
```

```{r multiple-path-summary}
fit_summary <- lapply(seq_along(fit_list), function(ii) {
  fit <- fit_list[[ii]]
  seg <- fit$segments_inferred
  data.frame(
    index_path = path_ids[ii],
    n_segments = nrow(seg),
    total_duration = sum(seg$durations),
    mean_speed = mean(seg$speeds),
    max_speed = max(seg$speeds),
    prop_motile_time = sum(seg$durations[seg$states == 1]) / sum(seg$durations)
  )
})
do.call(rbind, fit_summary)
```

## Next steps

* See [`CPLASS()`], [`CPLASS_multistart()`], and [`compute_csa()`] for full
  argument documentation.
* The companion paper (Do et al. 2026) documents the statistical model, the
  four Metropolis-Hastings proposal types, and the speed penalty in detail.

## References

Do, L., Do, D., Cook, K. J., Shen, Y., & McKinley, S. A. (2026).
Change-in-velocity detection in multidimensional data.

Cook, K. J., Rayens, N., Do, L., Payne, C. K., & McKinley, S. A. (2025).
Considering experimental frame rates and robust segmentation analysis of
piecewise-linear microparticle trajectories. *Mathematical Biosciences and
Engineering*, 22(10), 2595-2626.

Rayens, N. T., Cook, K. J., McKinley, S. A., & Payne, C. K. (2022).
Transport of lysosomes decreases in the perinuclear region: Insights from
changepoint analysis. *Biophysical Journal*, 121(7), 1205-1218.
